---
title: "Doublet Detection"
output:
  html_document: default
  pdf_document: 
    latex_engine: xelatex
date: "2024-11-12"
author: "Lauren Rylaarsdam"
---

Multiplets occur when two or more cells receive the same molecular identifier. High doublet rates can
complicate analysis as they dilute cell type-specific methylation patterns. We have developed an approach
for doublet removal by first simulating artificial doublets, constructing a Random Forest-based model to 
predict doublets, and removing cells from the original object exceeding the doublet score threshold.

The following vignette demonstrates how to apply this doublet removal approach using the brain vignette data.

########################################################################################################
## Loading libraries and example data

If you haven't already, load necessary libraries.

```{r, message=FALSE, warning=FALSE}
library(amethyst)
library(data.table)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tibble)
library(tidyr)
library(plyr)
library(cowplot)
library(harmony)
```

Next, download the example brain vignette data (this only needs to be done once; ignore if you went through the vignette).

```{r, eval=FALSE}
download.file("https://adeylabopen.s3.us-west-2.amazonaws.com/amethyst/brain_vignette_workspace.RData", "~/Downloads/brain_vignette_workspace.RData", method = "curl") 
download.file("https://raw.githubusercontent.com/lrylaarsdam/amethyst/main/vignettes/brain_vignette/brain_vignette_cellInfo.txt", "~/Downloads/brain_vignette_cellInfo.txt")
download.file("https://raw.githubusercontent.com/lrylaarsdam/amethyst/main/vignettes/brain_vignette/brain_vignette.annot", "~/Downloads/brain_vignette.annot", method = "curl")
```

Build the object according to steps in the brain vignette to obtain cell type annotations. 
If you have already followed the brain vignette, this step can be skipped if the data is still in your work space.

```{r, fig.width=4, fig.height=3, message=FALSE}
load("~/Downloads/brain_vignette_workspace.RData")
obj <- addCellInfo(obj, file = "~/Downloads/brain_vignette_cellInfo.txt")
obj <- addAnnot(obj, file = "~/Downloads/brain_vignette.annot", name = "method") 
obj@h5paths <- data.frame(row.names = rownames(obj@metadata), 
                          paths = rep("~/Downloads/brain_vignette.h5", length(rownames(obj@metadata))))
obj@metadata <- obj@metadata |> dplyr::filter(cov > 1000000 & mch_pct < 12)
set.seed(111)
obj@reductions[["irlba"]] <- runIrlba(obj, genomeMatrices = c("ch_100k_pct", "cg_100k_score"), dims = c(5, 5), replaceNA = c("mch_pct", 0))
set.seed(111)
obj@reductions[["irlba_regressed"]] <- regressCovBias(obj, reduction = "irlba")
set.seed(111)
obj <- runCluster(obj, k_phenograph = 10, reduction = "irlba_regressed") # consider increasing k_phenograph to 50 for larger datasets
set.seed(111)
obj <- runUmap(obj, neighbors = 5, dist = 0.01, method = "euclidean", reduction = "irlba_regressed") 
obj@metadata[["type"]] <- dplyr::recode(obj@metadata[["cluster_id"]],
                                        "1" = "Astro", 
                                        "2" = "Exc",
                                        "3" = "Micro",
                                        "4" = "Oligo",
                                        "5" = "Inh")

# define colors 
pal <- makePalette(option = 13, n = 5) 

# plot
dimFeature(obj, colorBy = type, colors = pal, pointSize = 1) +
  geom_text_repel(aes(label = type), color = "black", data = obj@metadata |> 
                    dplyr::group_by(type) |> 
                    dplyr::summarise(umap_x = median(umap_x), umap_y = median(umap_y)))
```

########################################################################################################
## Doublet detection

First, we simulate a fraction of artificial doublets. Note that you may have to downsample large datasets
to build the doublet model.

```{r}
# doublet calling
set.seed(123)
dbobj <- makeDoubletObject(obj, simFraction=0.30, threads = 1, genomeMatrices=c("cg_100k_score"))
```

Next, run IRLBA on the doublet-containing object.

```{r, fig.width=4, fig.height=3}
set.seed(123)
dbobj@reductions[["irlba"]] <- runIrlba(dbobj, genomeMatrices=c("cg_100k_score"), dims=c(5))
dbobj <- runCluster(dbobj, k_phenograph = 10, reduction = "irlba") # consider increasing k_phenograph to 50 for larger datasets
dbobj <- runUmap(dbobj, neighbors = 5, dist = 0.01, method = "euclidean", reduction = "irlba") 
dimFeature(dbobj, colorBy = doublet_info, colors = c("#299388", "#F29C5F"), pointSize = 1)
```

Use the object containing simulated doublets to build a prediction model.

```{r}
set.seed(123)
result <- buildDoubletModel(dbobj = dbobj, method="rf", reduction = "irlba")
```

Use the module to predict doublet scores.

```{r, fig.width=4, fig.height=3}
set.seed(123)
dbobj <- predictDoubletScores(dbobj = dbobj, model = result$model, reduction = "irlba")
ggplot(dbobj@metadata, aes(x = doublet_info, y = doublet_score, color = doublet_info, fill = doublet_info)) + 
  geom_boxplot(alpha = 0.1, outlier.size = 0, outlier.alpha = 0, width = 0.5) + geom_jitter(width = 0.1, size = 0.5) + theme_classic()
```

Notice the two artificial doublet outliers with a low doublet score. 
These doublets likely happened to be simulated from cells of the same type. 

Based on this graph, an appropriate threshold is around 0.3. 
Next we add the doublet scores to the original object and filter cells with a value exceeding our threshold.

```{r}
obj <- addDoubletScores(obj, dbobj)

singlets <- rownames(obj@metadata[obj@metadata$doublet_score < 0.3, ])
singlets <- subsetObject(obj, cells = singlets)
```

Now re-do clustering without the doublets.

```{r, fig.width=4, fig.height=3}
set.seed(123)
singlets@reductions[["irlba"]] <- runIrlba(singlets, genomeMatrices=c("cg_100k_score"), dims=c(5))
singlets <- runCluster(singlets, k_phenograph = 10, reduction = "irlba") # consider increasing k_phenograph to 50 for larger datasets
singlets <- runUmap(singlets, neighbors = 5, dist = 0.01, method = "euclidean", reduction = "irlba") 
dimFeature(singlets, colorBy = type, colors = pal, pointSize = 1) +
  geom_text_repel(aes(label = type), color = "black", data = singlets@metadata |> 
                    dplyr::group_by(type) |> 
                    dplyr::summarise(umap_x = median(umap_x), umap_y = median(umap_y)))
```

Compare the clustering results from the singlets object compared to the object with 
30% simulated doublets using the code below. Notice the clusters without doublets are tighter,
and we even seem to have resolved an additional group of inhibitory neurons.

```{r,  fig.width=8, fig.height=3}
plot_grid(dimFeature(dbobj, colorBy = doublet_info, colors = c("#264653", "#E76F51"), pointSize = 1) + ggtitle("with doublets"),
          dimFeature(singlets, colorBy = cluster_id, colors = c("#264653", "#2A9D8F", "#E9C46A", "#F4A261", "#E76F51"), pointSize = 1) + ggtitle("without doublets"))
```

We hope this helps! Feel free to reach out with any questions.

